---
title: "bellabeat"
author: "Folami"
date: "8/23/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## FitBit Fitness Tracker Data
### Data wrangling

The project involves analyzing the usage of smart devices in order to direct market strategy. Data for this exercise was obtained from Kaggle: [FitBit Fitness Tracker.](https://www.kaggle.com/datasets/arashnic/fitbit)

Following the download of the files, libraries were loaded
```{r Tidyverse}
#load libraries

library(tidyverse)
library(janitor)

```

The *dailyActivity_merged* and *sleepday_merged* table were imported using the *readr* package and examined.

```{r Import the tables}
#import tables and check the tables

daily_activity <- read_csv("dailyActivity_merged.csv")
glimpse(daily_activity)
str(daily_activity)

sleep_day <- read_csv("sleepDay_merged.csv")
head(sleep_day)
```

The *ActivityDate* variable (2nd column) of the *daily_activity* was then formatted from string  to date ("double").

```{r changes the Activitydate data type and save as date format}

#change the format and convert to date type
as.Date(daily_activity$ActivityDate, format= "%m/%d/%Y") -> activitydate
head(activitydate)

# to check the data_type
typeof(activitydate)

#replace the ActivityDate column with the new date format
daily_activity$ActivityDate <- activitydate
head(daily_activity)

```

The *sleepDay* column of the *sleep_day* was also formated

```{r format the SleepDay column}
sleepday <- as.POSIXct(sleep_day$SleepDay, format= "%m/%d/%Y %H:%M:%S")
sleep_day$SleepDay <- as.Date(sleepday)
head(sleep_day)
```

The distinct number of Id of the four tables where then checked to get an idea of which tables had less than the sampled number of users.

```{r obtain the distinct number of Id present in each table}

n_distinct(daily_activity$Id)
n_distinct(sleep_day$Id)

```

The tables were then joined using the full join. The tables were joined on the id and date variables of the respective tables. a new table: *merged_dataset* was formed.

```{r joining the tables}

merged_dataset <- full_join(daily_activity, sleep_day, by = c("Id" = "Id", "ActivityDate" = "SleepDay"))

```

After merging the data, further cleaning was done. Id was converted to character data type from number. 

```{r change the Id data_type and remove duplicates}
#change the Id data_type and remove duplicates.

merged_dataset$Id <- as.character(merged_dataset$Id)

merged_dataset <- distinct(merged_dataset)

```

The column names were cleaned thereafter

```{r cleaning column names}

merged_dataset <- clean_names(merged_dataset)
merged_dataset <- rename(merged_dataset, "date" = activity_date)

```

then the columns intended for analysis were selected.

```{r select columns for analysis}

merged_dataset <- merged_dataset %>%  select(id : total_distance, very_active_distance : total_time_in_bed)

```

The final table was examined using the following before exploratory visualization was done

```{r}
#checkout merged_dataset
View(merged_dataset)
colnames(merged_dataset)
str(merged_dataset)
n_distinct(merged_dataset$id)

```

### Exploratory visualizations

Some graphs were plotted to check for relationships between the different variables

```{r relationship between calories and total_steps}

ggplot(data = merged_dataset) + geom_point(aes(x = total_steps , y = calories, color = id)) + labs(x="Total steps", y = "Calories", title="Calories vs. Total_steps")

```


Visual shows a positive correlation between the variables.  

```{r #export table}

write.csv(merged_dataset, "merged_dataset.csv")

```
The cleaned dataset (*merged_dataset*) is exported, further explored and visualized using a BI tool and to draw further insights from the dataset.
---
